---
title: "Mixed Linear Effects Modeling"
subtitle: "Workshop on Statistics for Linguistics"
author: "Sky Onosson, University of Manitoba"
output: 
  html_document:
    number_sections: true
---

<big>

```{r, include=FALSE, message=F, warning=F}
library("knitr")
opts_chunk$set(echo = TRUE)
```

---

## Agenda {-}

* Mixed-Effects Modeling

---

## Load required packages and LIPP data {-}

```{r, message=F, warning=F}
if(!require(tidyverse)){install.packages("tidyverse")}
library(tidyverse)

if(!require(viridis)){install.packages("viridis")}
library(viridis)

LIPP <- read_csv("LIPP.csv")
```

# Mixed-Effects Modeling

<http://www.bodowinter.com/uploads/1/2/9/3/129362560/bw_lme_tutorial1.pdf>

<http://www.bodowinter.com/uploads/1/2/9/3/129362560/bw_lme_tutorial2.pdf>

One of the most common methods of statistical analysis in fields like sociolinguistics is mixed-effects (regression) modeling (MEM). As the full name implies, this is a type of regression model similar to linear regression, but with the addition of "mixed effects". As noted by Sonderegger et al., MEMs offer a range of benefits:

> * More than one grouping unit can be included in the same model
* Unequal number of observations per level is OK
* **Explicitly model variability** among levels of the same grouping factor
    * Ex: By participant variability ("individual differences")
    * This means the analyst can explore and test questions about the variability itself (as opposed to effects averaging across participants, etc.), which is not possible using RM-ANOVA, or other methods we've covered so far.

The suitability of MEMs for statistical analysis of data from variationist sociolinguistics and related fields should be quite apparent given these points.











</big>